\section{Experiments}
\label{sec:experiments}

We evaluate \splitzip around five questions. (RQ1) Does it reduce transfer time in a real PD transfer stack? (RQ2) Does the BF16 path reconstruct KV caches exactly? (RQ3) Which algorithmic components matter? (RQ4) How does it compare with raw BF16, native FP8, and other compression methods? (RQ5) Does the exponent pattern hold across models?

\subsection{Setup}

\textbf{Platform and timing protocol.}
All codec experiments are conducted on NVIDIA H200 GPUs. We report compression and decompression throughput in GB/s, where the byte count is the uncompressed native tensor size. Each codec benchmark first verifies bitwise round-trip correctness and then times repeated GPU executions after warm-up. For the baseline comparison, all methods operate on the same real-activation BF16 tensor of shape $32768 \times 4096$. For transfer studies, we compose measured \splitzip encode/decode time, measured GPU host-device DMA time, and a staged layer-pipeline transfer model. The modeled transport modes are CPU-RDMA at 47~GB/s and RoCE 4$\times$200G at 87~GB/s effective payload bandwidth. Therefore, the sequence-length transfer curves are projections from measured codec/DMA components, not direct multi-node RoCE measurements. The Qwen3-32B breakdown uses additive sequential accounting, so encode, transfer, and decode components sum exactly in the stacked-bar visualization.

\textbf{Datasets and workloads.}
Transfer workloads use authentic BF16 KV blocks extracted from one 1024-token forward pass per model and then assembled to target sequence lengths. We evaluate Llama-3-8B and Qwen3-30B-A3B from 512 to 65,536 tokens under both transport modes. For Qwen3-32B, we measure the transmission breakdown at sequence lengths 2,048, 8,192, 16,384, 32,768, and 65,536, and plot the 2,048, 16,384, and 65,536 cases. FP8 experiments also use Qwen3-32B and evaluate exact compression of native E4M3 and E5M2 tensors from 1,024 to 65,536 tokens. Calibration-transfer experiments calibrate the BF16 top-16 codebook on WikiText-2 train and evaluate coverage on WikiText-2 test, HumanEval, GSM8K, MMLU, and Penn Treebank.

\textbf{Baselines and ablations.}
For BF16 codec comparison, we compare nvCOMP LZ4, DFloat11, ZipServ, ZipNN, and \splitzip. nvCOMP, DFloat11, ZipServ, and \splitzip are measured locally on the same $32768 \times 4096$ tensor; ZipNN uses the reported 66\% compressed-size fraction and reported 1.15/1.65~GB/s compression/decompression throughput. For transfer-time studies, the native baseline moves raw BF16 or FP8 cache bytes, while \splitzip performs encode, compressed transfer, and decode. We further ablate BF16 top-8 3-bit coding versus top-16 4-bit coding, explicit escape-position storage versus sentinel-only escape discovery, pre-calibrated versus dynamic top-16 codebooks, and per-tensor, per-token, and per-channel calibration granularity.

\subsection{Real BF16 KV transfer through Mooncake}
\label{sec:real-bf16}

Table~\ref{tab:real-kv-mooncake} reports transfer measurements. Controlled Mooncake tests show that compressed transfers are faster than raw BF16 across tested sizes. The real Qwen3-30B-A3B row is the most representative setting: on 402.7~MB of BF16 KV data, \splitzip reduces the transfer-only median from 146.88~ms to 115.44~ms and improves the full encode-transfer-decode path by $1.238\times$ while preserving bitwise correctness.

\input{tables/real_kv_mooncake}

The real-KV compression ratio is lower than the controlled $1.333\times$ value because Qwen3-MoE has higher exponent entropy and a higher escape rate. This is the expected tradeoff of a lossless escape stream: the method keeps exactness, but realistic rare exponents consume side-stream bytes.

\subsection{Algorithm ablation}
\label{sec:ablation}

Table~\ref{tab:codec-ablation} ablates the main design choices. Splitting sign+mantissa from exponent is not enough by itself; compression comes from coding the exponent stream. Removing the escape stream gives the nominal 4-bit payload but loses exactness on rare exponents. The full BF16 codec restores those exponents and is bitwise exact. The 3-bit variant gives a higher ratio by keeping only top-8 exponent codes, but it is a near-lossless mode and is therefore secondary to the lossless BF16 contribution.

\input{tables/codec_ablation}

\subsection{Native BF16 and native FP8 precision}
\label{sec:native-precision}

Table~\ref{tab:native-precision} separates two notions that are easy to conflate. BF16+\splitzip is lossless with respect to BF16. FP8+\splitzip is lossless with respect to native FP8 tensors, but it inherits the quantization error of FP8 relative to BF16. We obtain the FP8 profile by converting measured BF16 KV tensors to FP8 layouts and then compressing the FP8 byte representation. In deployments already serving native FP8 KV caches, the current exact compact codecs provide additional lossless savings of $1.059\times$ on E4M3 and $1.214\times$ on E5M2. The entropy profile suggests larger possible ratios, but Table~\ref{tab:native-precision} reports the implemented lossless GPU path.

\input{tables/native_precision}

Table~\ref{tab:format-transfer-latency} composes these implemented ratios with measured Mooncake TCP loopback transfer times and GPU encode/decode throughput. On this local loopback setup, BF16+\splitzip improves full-path transfer latency by $1.225\times$. E4M3+\splitzip is slightly slower than raw native E4M3 because its $1.059\times$ ratio is too small to cover encode/decode overhead at this loopback bandwidth, although it remains $1.840\times$ faster than raw BF16 transfer. E5M2+\splitzip gives a useful native-FP8 speedup of $1.111\times$ and a $2.200\times$ speedup relative to raw BF16 transfer.

\input{tables/format_transfer_latency}

\subsection{Layer-pipeline projection}

Table~\ref{tab:pipeline} uses measured controlled-kernel timings to project layer-pipelined transfer for a Llama-3-70B-style 64K-context request. The lossless BF16 mode closely tracks the $1.333\times$ controlled compression ratio at 15--87~GB/s. At very high link bandwidth, the margin depends on maintaining optimized encode throughput; the real-KV result in Table~\ref{tab:real-kv-mooncake} is therefore the more conservative serving-path evidence.

\input{tables/pipeline_speedup}

\subsection{Correctness and generation quality}

Table~\ref{tab:correctness} verifies exact reconstruction across context lengths. In lossless BF16 mode, all tested contexts have zero reconstruction errors and zero logit difference. Across 140 generation prompts, Qwen2.5-1.5B and TinyLlama-1.1B produce identical text under uncompressed BF16 and lossless \splitzip reconstruction.

\input{tables/correctness_quality}

\subsection{Comparison with other methods}

Table~\ref{tab:baselines} positions \splitzip against raw transfer, general-purpose lossless codecs, native FP8, and related neural-network compression systems. FP8 gives a larger nominal size reduction but is lossy relative to BF16. CPU lossless codecs are either ineffective on BF16 KV bytes or too slow for a latency-sensitive transfer path. \splitzip occupies a different point: exact BF16 reconstruction, GPU execution, and integration with the PD transfer path.

\input{tables/baseline_comparison}

\subsection{Limitations}
\label{sec:limitations}

\splitzip is most useful when KV transfer is on the TTFT critical path. It does not help monolithic serving where KV tensors are never transferred across workers. The BF16 lossless ratio is modest compared with lossy FP8, and real models with more diverse exponent distributions reduce the ratio through larger escape streams. Our real Mooncake measurements use TCP transport on a single-node setup; RDMA should benefit from reduced byte count, but absolute speedups need direct RDMA validation. Finally, the optional 3-bit variant is included as an ablation and secondary operating point, not as the primary correctness guarantee.

\textbf{Broader impact.}
Reducing KV-transfer overhead can lower the cost and energy footprint of LLM serving. As with other inference-efficiency improvements, it can also lower the cost of serving harmful or low-quality models; \splitzip changes the serving system's data movement, not model capabilities or safety behavior.
