\section{Our Method}
\label{sec:method}

\subsection{KV transfer setting}

In PD serving, the prefill worker produces one key tensor and one value tensor per layer. Before decode can start, these tensors must be made available on the decode worker. A layer with $n$ BF16 KV elements occupies $2n$ bytes, and total transfer volume scales linearly with sequence length and model depth. \splitzip is inserted between KV generation and the transfer call: the prefill worker encodes each layer's KV tensor, sends the compressed payload, and the decode worker reconstructs the original tensor before attention uses it.

\subsection{Exponent concentration in BF16 KV caches}
\label{sec:exponent}

BF16 values consist of one sign bit, eight exponent bits, and seven mantissa bits. Directly compressing the raw 16-bit words is ineffective because sign and mantissa bits are high-entropy. The exponent field, however, is strongly concentrated. Table~\ref{tab:exponent} summarizes the profiling evidence: dense Qwen, TinyLlama, and Phi models all have top-15 exponent coverage above 99.5\%, and the real Qwen3-30B-A3B MoE run remains above 99.0\% despite a higher escape rate.

\input{tables/exponent_statistics}

This pattern motivates a split representation. \splitzip stores the sign+mantissa byte unchanged and encodes only the exponent byte. The result is a codec with predictable output size, simple GPU kernels, and no entropy-coding serial dependency.

\subsection{Lossless BF16 codec}
\label{sec:bf16-codec}

For each BF16 word, \splitzip forms two byte streams:
\begin{itemize}[nosep,leftmargin=*]
\item \textbf{Exponent byte:} bits $[14:7]$, the biased BF16 exponent.
\item \textbf{Sign+mantissa byte:} the sign bit and seven mantissa bits, stored verbatim.
\end{itemize}

The BF16 lossless mode builds a 256-entry encode lookup table per calibrated model or layer. The 15 most frequent exponents map to codes 0--14, while all remaining exponents map to code 15, the escape code. The encoder packs two 4-bit codes into one byte and appends the sign+mantissa stream. For every escaped element, it records the element position and the original exponent value. Decoding unpacks nibbles, reconstructs common exponents from the 16-entry decode table, restores the raw sign+mantissa bits, and patches escaped positions.

For any BF16 KV tensor $\mathbf{x}$, the lossless mode satisfies
\[
  \mathrm{decode}(\mathrm{encode}(\mathbf{x})) = \mathbf{x}
  \quad\text{bitwise.}
\]
The nominal compressed size is $n/2+n+\epsilon$ bytes for $n$ BF16 elements, where $\epsilon$ is the escape-stream overhead. This gives $1.333\times$ compression when escapes are negligible; on real Qwen3-30B-A3B KV caches, the measured ratio is $1.293$--$1.301\times$ because the escape rate is higher.

\subsection{Native FP8 extension}
\label{sec:fp8-codec}

\splitzip is not a replacement for FP8 quantization. Instead, it is a lossless codec for whichever native KV representation a serving system already uses. For native FP8 caches, the same decomposition applies to FP8 E4M3 and E5M2 words: sign and mantissa bits are stored verbatim, while exponent bits are coded. This gives additional compression on top of the $2\times$ storage reduction of FP8, but exactness is with respect to the FP8 tensor, not the original BF16 tensor.

\subsection{GPU implementation and Mooncake integration}

The codec is implemented with Triton kernels. The BF16 encode kernel loads BF16 words as 16-bit integers, extracts exponents, looks up 4-bit codes, packs nibbles, and writes sign+mantissa bytes. The decode kernel reverses this flow and then applies a small escape-fix kernel. The optimized controlled H200 microbenchmark reaches 252~GB/s encode and 1460~GB/s decode for a 268~MB layer-scale tensor.

We integrate \splitzip with Mooncake Transfer Engine using its managed-buffer API. On the prefill side, the compressed payload contains packed exponents, sign+mantissa bytes, escape positions, escape values, and a compact metadata header. Mooncake transfers this smaller buffer unchanged. On the decode side, \splitzip reconstructs the BF16 tensor on GPU before decode attention consumes it.

\subsection{Layer-pipelined execution}

A serial implementation would encode all layers, transfer all compressed bytes, and then decode all layers. \splitzip instead pipelines layers:
\[
T_{\text{pipeline}} =
T_{\text{enc}}^{(1)} +
N \cdot \max(T_{\text{enc}},T_{\text{xfer}},T_{\text{dec}}) +
T_{\text{dec}}^{(N)},
\]
where $N$ is the number of layers. When network transfer is the bottleneck, the pipeline speed approaches the compression ratio. When links become very fast, encode throughput and escape handling can dominate; Section~\ref{sec:experiments} reports both controlled and real-KV measurements to make this distinction explicit.
