\section{Related Work}
\label{sec:related}

\textbf{Disaggregated PD serving.}
Mooncake~\cite{mooncake} introduces a KV-cache-centric disaggregated architecture with a transfer engine supporting RDMA and TCP.
DistServe~\cite{distserve} separates prefill and decode with pull-based KV retrieval, reporting $<$0.1\% overhead from KV transfer.
Splitwise~\cite{splitwise} pipelines layer-wise transfers via NCCL for heterogeneous GPU setups.
FlowKV~\cite{flowkv} addresses KV memory fragmentation with segmented kernel launches, achieving 96\% latency reduction over vLLM disaggregated mode.
None of these systems compress KV data during transfer.

\textbf{Lossless weight compression.}
DFloat11~\cite{dfloat11} applies Huffman coding to BF16 weight exponents, achieving $\sim$30\% compression at 584~GB/s GPU decode.
ZipServ~\cite{zipserv} introduces Tensor-Core-Aware Triple Bitmap Encoding fused with GEMM.
ZipNN~\cite{zipnn} extends exponent-mantissa Huffman coding to FP8 and KV cache tensors in a storage context.
These methods target weight storage and loading, not the KV transfer path in serving.

\textbf{Lossy KV compression.}
KIVI~\cite{kivi} and KVQuant apply per-channel or per-token quantization to KV caches for memory reduction during serving.
HACK~\cite{hack} enables homomorphic attention computation on quantized KV data for disaggregated serving, avoiding dequantization overhead.
These approaches sacrifice bit-exactness, which SplitZip avoids.

\textbf{Communication compression.}
MegaScale-MoE~\cite{megascalemoe} applies FP8 compression to MoE all-to-all communication during training.
Flash Communication~\cite{flashcomm} uses low-bit compression for tensor-parallel inference.
These target training gradients or TP activations, not PD KV transfer.

\textbf{GPU lossless compression.}
NVIDIA nvCOMP provides GPU-accelerated LZ4, Cascaded, and ANS codecs.
In our experiments, LZ4 achieves zero compression on BF16 KV data because it cannot exploit floating-point exponent structure.
SplitZip fills this gap with a format-aware codec.
