\section{Related Work}
\label{sec:related}

\textbf{Disaggregated PD serving.}
Mooncake~\cite{qin2024mooncake} introduces a KV-cache-centric disaggregated architecture with a Transfer Engine for RDMA and TCP transport.
DistServe~\cite{distserve} separates prefill and decode to improve goodput under latency constraints, while Splitwise~\cite{splitwise} pipelines phase-split inference across heterogeneous GPUs.
FlowKV~\cite{flowkv} focuses on fragmented KV memory transfer and segmented kernel launches.
These systems reduce scheduling or transfer overhead, but they do not make KV transfer itself smaller through a lossless, GPU-native codec.

\textbf{Lossless LLMs compression.}
DFloat11~\cite{dfloat11} compresses BF16 model weights by coding exponent bits while preserving sign and mantissa information. ZipNN~\cite{zipnn} studies lossless compression for weights, checkpoints, and low-precision tensors. ZipLLM~\cite{zipllm} is also related, as it combines tensor-level deduplication with lossless compression for efficient LLM storage.
\splitzip is closest in spirit to this line of work, but its target is different: the latency-critical KV transfer path in disaggregated serving, where compression, network transfer, and decompression must be overlapped at layer granularity.

\textbf{Communication compression.}
MegaScale-MoE~\cite{megascalemoe} uses FP8 compression for MoE communication during training, and Flash Communication~\cite{flashcomm} targets tensor-parallel inference communication.
Those approaches reduce distributed tensor traffic, but they address training or tensor-parallel activation exchange rather than PD KV-cache handoff.
\splitzip focuses on the serialized KV payload that sits directly on the TTFT path.

\textbf{GPU lossless codecs.}
General GPU compression libraries such as nvCOMP provide LZ4\cite{nvidia2026nvcomp}, Cascaded, and entropy-coding primitives.
They are valuable for generic data, but floating-point KV tensors need a format-aware decomposition: the BF16 mantissa stream is hard to compress, while the exponent stream is highly structured.
\splitzip uses this decomposition directly instead of applying byte-stream compression to the whole tensor.
